---
title: "Flip: Data analysis"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = TRUE
)
options(knitr.kable.NA = '-')

# load packages
library(tidyverse)    # tidyr packages
library(lmSupport)    # for linear model analysis
library(lme4)         # for linear mixed effects models (LMM)
library(car)          # for ANOVAs (computing Wald's p-values)
library(predictmeans) # for checking LMM assumptions
library(emmeans)      # for retrieving estimates
library(psych)
library(knitr)        # for formatting tables
library(broom)        # for tidy lm output
library(broom.mixed)  # for tidy lmer output
library(lattice)      # for zero margins in plots
library(patchwork)    # for grouping plots
```

First, we read in the data and compute both the novelty preference (total looking time to the novel stimuli minus total looking time to familiar stimuli), as well as the novelty preference **corrected for total looking time**, i.e. the "novelty preference score" described in the summary from 10 July 2019.

```{r import_data, message=TRUE, warning=FALSE, include=FALSE, paged.print=FALSE}

data <- read.csv("../Data/flip.csv") %>%                  # read in data
  mutate(
    novel_pref       = Novel-Familiar,                    # uncorrected difference score
    novel_pref_corr  = (Novel-Familiar)/(Familiar+Novel), # corrected difference score
    avg_looking_time = (Novel+Familiar)/2                 # mean looking time
)
```

## Compare distributions of Syntax and Saffran & Wilson

The first thing I think that it makes sense to do is just to look at the distribution of (average) looking time for familiar and novel stimuli in the two studies. This seems like a good first step before combining the data from the two studies.

```{r format_data, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# convert data to long format
# also perform some centering transformations that will be useful in fitting models
data_long <- gather(data, "trialType", "lookingTime", Familiar, Novel) %>%
  mutate(
    trialType_center = case_when(
      trialType == "Familiar" ~ -0.5,
      trialType == "Novel"    ~ 0.5),
      HPP = as.factor(HPP),
      HPP_center = HPP - mean(HPP)
  )
```

```{r plot_timeTypeStudyLocation, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

ggplot(data_long, aes(trialType, lookingTime, fill = Study))+
  facet_wrap(~ Study + Location) +
  geom_violin(size = 1, alpha = 0.5, colour = NA) +
  geom_point(width = 0.01, size = 2, alpha = 0.5)+
  geom_line(aes(group = Participant), color = "black", alpha = 0.2, size = 0.25) +
  stat_summary(fun.data = mean_se, geom = "crossbar", colour = "black", size = 0.75, fill = "transparent", width = 0.5) +
  labs(x = "Trial type", y = "Looking time (ms)") +
    scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none" 
  ) +
  ggsave("../Figures/lookingTimes.png")

```

Looking at the average looking times, it seems from visual inspection that the distributions look somewhat similar. Perhaps the looking times for Syntax in Wisconsin are somewhat shorter, mainly because the looking times to familiar items look substantially shorter in Wisconsin compared to Barcelona. 

In general, there's nothing that looks too dramatic here to me suggesting that one distribution looks radically different than the others (in shape or in the range of values).

## Linear (Correlational) Analyses

As a first step, I'm just going to look at the linear relationship between visit number and headturn preference, not accounting for non-independence due to other factors (e.g., study). This helps for getting a sense of the simple effect and how robust it is to different sources of non-independence/ analytical decisions. I'll also look at how much the effect relies on high leverage points (children with higher visit numbers) and visualize the effect for each study and location.

### Visit number predicts magnitude of novelty/ familiarity preference {.tabset}

#### Uncorrected novelty preference

Fit a linear model predicting novelty preference from number of HPP visits (equivalent to a correlation).

```{r model1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
# novelty preference
model1 <- lm(novel_pref ~ HPP, data = data)

# model summary
model1 %>%
  summary() %>%
  tidy() %>%
  mutate(term = c("Intercept", "HPP")) %>%
  kable(digits = 3, col.names = c("", "Coefficient", "SEM", "t", "p"), align = "c")
```

```{r plot_model1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

ggplot(data, aes(HPP, novel_pref, colour = Study)) +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model1)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(x = "Number of HPP visits", y = "Novelty Preference (ms)\n(not corrected for total looking time)") +
  scale_colour_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none" 
  ) +
    ggsave("../Figures/model1.png")
```

#### Corrected novelty preference

Fit a linear model predicting novelty preference from number of HPP visits (equivalent to a correlation).

```{r model2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}

# corrected novelty preference
model2 <- lm(novel_pref_corr ~ HPP, data = data)

# model summary
model2 %>%
  summary() %>%
  tidy() %>%
  mutate(term = c("Intercept", "HPP")) %>%
  kable(digits = 3, col.names = c("", "Coefficient", "SEM", "t", "p"), align = "c")
```

```{r plot_model2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

ggplot(data, aes(HPP, novel_pref_corr, colour = Study)) +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model2)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_colour_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none" 
  ) +
  ggsave("../Figures/model2.png")
```


### Split the prediction by study and location {.tabset}

#### Split by study (Santolin vs. Saffran & Wilson)

```{r split_study, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# split by study

ggplot(data, aes(HPP, novel_pref_corr, colour = Study)) +
  facet_wrap(~Study) + 
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.1)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model2)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0) +
  labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_colour_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none" 
  ) +
  ggsave("../Figures/model2.png")
```


#### Split by study and location (Wisconsin vs. Barcelona)

```{r plot_cnpHppStudy, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

ggplot(data, aes(HPP, novel_pref_corr, colour = Study)) +
  facet_wrap(~ Study + Location) +
  geom_point(size = 3, alpha = 0.5, position = position_jitter(width = 0.01)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model2)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_colour_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
  ) +
  ggsave("../Figures/model2_location.png")
```

### How robust is this analysis to reduced numbers of visits {.tabset}

#### 1 - 5 visits

Fit a linear model predicting novelty preference from number of HPP visits (equivalent to a correlation).

```{r model3, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
# novelty preference
model3 <- lm(novel_pref_corr ~ HPP, data = data)
#modelSummary(m)
model3 %>%
  summary() %>%
  tidy() %>%
  mutate(term = c("Intercept", "HPP")) %>%
  kable(digits = 3, col.names = c("", "Coefficient", "SEM", "t", "p"), align  = "c")
```

```{r plot_model3, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

ggplot(data, aes(HPP, novel_pref_corr, color = Study)) +
  geom_point(size = 3, alpha = 0.7, position = position_jitter(width = 0.05)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model3)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
    labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_colour_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5)) +
theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
  ) +
  ggsave("../Figures/model3.png")
```

#### 1 - 4 visits

Fit a linear model predicting novelty preference from number of HPP visits (equivalent to a correlation).

```{r model4, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}

# novelty preference
model4 <- lm(novel_pref_corr ~ HPP, data = filter(data, HPP < 5))

# model summary
model4 %>%
  summary() %>%
  tidy() %>%
  mutate(term = c("Intercept", "HPP")) %>%
  kable(digits = 3, col.names = c("", "Coefficient", "SEM", "t", "p"), align = "c")
```

```{r plot_model4, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data %>%
  filter(HPP < 5) %>%
  ggplot(aes(HPP, novel_pref_corr, color = Study))+
  geom_point(size = 3, alpha = 0.7, position = position_jitter(width = 0.05)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model4)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
    labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_colour_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = c(1, 2, 3, 4)) +
theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
) +
  ggsave("../Figures/model4.png")
```

#### 1 - 3 visits

Fit a linear model predicting novelty preference from number of HPP visits (equivalent to a correlation).

```{r model5, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
# novelty preference
model5 <- lm(novel_pref_corr ~ HPP, data = filter(data, HPP < 4))

# model summary
model5 %>%
  summary() %>%
  tidy() %>%
  mutate(term = c("Intercept", "HPP")) %>%
  kable(digits = 3, col.names = c("", "Coefficient", "SEM", "t", "p"), align = "c")
```

```{r plot_model5, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data %>%
  filter(HPP < 4) %>%
  ggplot(aes(HPP, novel_pref_corr, color = Study))+
  geom_point(size = 3, alpha = 0.7, position = position_jitter(width = 0.05)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model5)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
    labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_x_continuous(breaks = c(1, 2, 3)) +
  scale_colour_brewer(palette = "Dark2") +
theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
) +
  ggsave("../Figures/model5.png")
```

#### 1 - 2 visits

Fit a linear model predicting novelty preference from number of HPP visits (equivalent to a correlation).

```{r model6, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
# novelty preference
model6 <- lm(novel_pref_corr ~ HPP, data = filter(data, HPP < 3))
# model summary
model6 %>%
  summary() %>%
  tidy() %>%
  mutate(term = c("Intercept", "HPP")) %>%
  kable(digits = 3, col.names = c("", "Coefficient", "SEM", "t", "p"), align = "c")
```

```{r plot_model6, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data %>%
  filter(HPP < 3) %>%
  ggplot(aes(HPP, novel_pref_corr, color = Study))+
  geom_point(size = 3, alpha = 0.7, position = position_jitter(width = 0.05)) +
  stat_summary(fun.data = mean_se, geom = "pointrange", colour = "black", size = 0.75) +
  stat_summary(aes(y = fitted(model6)), fun.y = mean, geom = "line", colour = "black", size = 1) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(x = "Number of HPP visits", y = "Novelty Preference (%)\n(corrected for total looking time)") +
  scale_x_continuous(breaks = c(1, 2)) +
  scale_colour_brewer(palette = "Dark2") +
theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
) +
  ggsave("../Figures/model6.png")
```

## Linear mixed-effects analyses

### Reproducing analysis with summarized difference scores

Here we specify a model that tries to predict the corrected novelty preference score (CNPS) as a function of the number of HPP, incluing a random intercept the variable *Study*. This essentially reproduces the model reported in the shared analysis script from
10 July 2019, with a few slight technical differences. For instance, I will simply fit the model with the maximal random effects structure here as the "default" and use Wald chi-squared tests to get p-values (not my preferred method, which is Kenward-Rogers approximantion, but this has some odd issues in this instance - see below).

```{r model7, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data7 <- data %>% mutate(HPP_center = HPP - mean(HPP)) # center hpp (for easier model fitting)

model7 <- lmer(novel_pref_corr ~ HPP_center + (1 + HPP_center | Study), data = data7, REML = FALSE)
broom.mixed::tidy(model7) %>% 
  select(term, group, effect, estimate, std.error) %>%
  left_join(., tidy(Anova(model7, type = "III", test.statistic = "Chisq")), by = "term") %>%
  mutate(
    term = c("Intercept", "HPP", "Intercept SD", "HPP SD", "HPP Cor Intercept", "Observation SD"),
    group = c(NA, NA, "Study", "Study", "Study", "Residual"),
    effect = c(rep("Fixed", 2), rep("Random", 4)),
    ) %>%
  kable(digits = 3, col.names = c("", "Group", "Type", "Estimate", "SE", "Chi^2", "DF", "P-value"), align = "c")
```

`lmer(novel_pref_corr ~ hpp_center + (1 + hpp_center | Study), data = data)`

$$ \begin{aligned}CNPS_{ps} = \beta_{0} + Study_{0s} + (\beta_{1} + Study_{1s}) HPP_{i} + e_{i}\end{aligned}$$

### Approach using long-format data & lme4

The best (most powerful) analysis would involve fitting linear mixed-effects models to the trial-by-trial looking time data (*LT*). This model would look something like this, fit on the trial-level looking data:

`lmer(lookingTime ~ trialType * hpp + (1 + trialType | baby) + (1 | item) + (1 + trialType * hpp | Study))`

$$ \begin{aligned}LT_{ips} = \beta_{0} + Item{0i} + Participant_{0p} + Study_{0s} + \\ (\beta_{1} + Participant_{1p}) HPP_{i} + (\beta_{2} + Study_{1s}) Condition_{i} + (\beta_{3} + Participant_{3p} + Study_{3s}) (Condition \times HPP)_{ips} + e_{ips} \end{aligned}$$

For now, let's consider a mixed-effects model fit to the average looking time data.

I think the model that we want to test would involve predicting an interaction between item type (familiar vs. novel) and number of hpp visits on looking time - the effect of familiar vs. novel items on looking behavior **depends on** the number of headturn visits an infant has had.

`lmer(lookingTime ~ trialType * hpp + (1 | baby) + (1 + hpp | Study))`

$$ \begin{aligned} LT_{is} = \beta_{0} + Participant_{0p} + Study_{0s} + \\ (\beta_{1} + Participant_{1p} + Study_{1s}) HPP_{i} + Condition_{c} + e_{is} \end{aligned}$$

We include a by-participant random intercept, since we have multiple observations per baby. Since item type has only two levels, we fit the model without a by-participant random slope (since this overdetermines the model; my understanding is that it's OK to still force the model to accept the random slope, but it shouldn't much matter for the estimates.

```{r model8, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

model8 <- lmer(lookingTime ~ trialType_center * HPP_center + (1 | Participant)  + (1 + HPP_center | Study), data = data_long, REML = FALSE)

broom.mixed::tidy(model8) %>% 
  select(term, group, effect, estimate, std.error) %>%
  left_join(., tidy(Anova(model8, type = "III", test.statistic= "Chisq")), by = "term") %>%
  mutate(
    term = c("Intercept", "Trial type", "HPP", "Trial type * HPP", "Intercept SD", "Intercept SD", "HPP SD", "HPP Cor Intercept", "Observation SD"),
    group = c(NA, NA, NA, NA, "Participant", "Study", "Study", "Study", "Residual"),
    effect = c(rep("Fixed", 4), rep("Random", 5))
    ) %>%
  kable(digits = 3, col.names = c("", "Group", "Type", "Estimate", "SE", "Chi^2", "DF", "p"), align = "c")

model8.summary <- summary(model8)

```

The significant interaction term suggests that the number of visits changes the relationship between item type and looking time, i.e. the nature of the familiarity/novelty preference.

The way to get appropriate p-values from lmer models is of course a matter of debate. Here, we used a Wald chi-squared test. Typically, the method I prefer is the Kenward-Rogers approximation which estimates the "correct" degrees of freedom. However, there's something a bit odd going on when adding the hpp by-study random slope with this method, which is probably because study only has two levels. The chi-squared method should be fine, but this is also something we could do a deeper dive on down the road (it might also be just another argument against including the hpp by-study random slope).


```{r plot_coefs8, echo=FALSE, message=FALSE, warning=FALSE}
broom.mixed::tidy(model8) %>% 
  filter(effect == 'fixed', term != "(Intercept)") %>%
  mutate(term = c("Trial type", "HPP", "Trial time * HPP")) %>%
  ggplot(aes(estimate, term, colour = term)) +
  geom_point(size = 5) +
  geom_errorbarh(aes(xmax = estimate+std.error, xmin = estimate-std.error), height = 0, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Estimate", y = "Term") +
  scale_colour_brewer(palette = "Dark2") +
theme(
    panel.grid.major.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
) +
  ggsave("../Figures/model8_coefs.png")
```

Now we interpret the interaction:

```{r plot_model8, echo=FALSE, message=FALSE, warning=FALSE}
data_summary <- data_long %>%
                  group_by(HPP, trialType) %>%
                  summarise(mean = mean(lookingTime))
data_long %>%
  mutate(trialType = factor(trialType, c("Familiar", "Novel"))) %>%
  ggplot(aes(x = trialType, y = lookingTime)) +
  facet_wrap(~HPP, nrow = 1) +
  zero_margin + 
  geom_violin(aes(fill = trialType), colour = NA, alpha = 0.5) +
  geom_line(aes(group = Participant), colour = "black", size = 0.25, alpha = 0.2) +
  geom_line(data = data_summary, aes(x = trialType, y = mean), size = 1, colour = "black") +
  stat_summary(fun.data = mean_se, geom = "pointrange", group = 1, size = 0.5) +
  stat_summary(fun.y = mean, geom = "line", group = 1, size = 0.5) +
  labs(x = "HPP", y = "Looking time (ms)") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
) +
  ggsave("../Figures/model8_interaction.png")

```

### Is the effect robust when we remove babies with high numbers of visits (4 or 5)?

Checking here that the effect still holds when we remove high leverage points - the handful of babies who had 4 or 5 visits.

```{r model9, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

model9 <- lmer(lookingTime ~ trialType_center * HPP_center + (1 | Participant) + (1 + HPP_center | Study), data = filter(data_long, HPP < 4), REML = FALSE)

broom.mixed::tidy(model9) %>% 
  select(term, group, effect, estimate, std.error) %>%
  left_join(., tidy(Anova(model8, type = "III", test.statistic= "Chisq")), by = "term") %>%
  mutate(
    term = c("Intercept", "Trial type", "HPP", "Trial type * HPP", "Intercept SD", "Intercept SD", "HPP SD", "HPP Cor Intercept", "Observation SD"),
    group = c(NA, NA, NA, NA, "Participant", "Study", "Study", "Study", "Residual"),
    effect = c(rep("Fixed", 4), rep("Random", 5))
    ) %>%
  kable(digits = 3, col.names = c("", "Group", "Type", "Estimate", "SE", "Chi^2", "df", "p"), align = "c")

```

```{r plot_coefs9, echo=FALSE, message=FALSE, warning=FALSE}
broom.mixed::tidy(model9) %>% 
  filter(effect == 'fixed', term != "(Intercept)") %>%
  mutate(term = c("Trial type", "HPP", "Trial time * HPP")) %>%
  ggplot(aes(estimate, term, colour = term)) +
  geom_point(size = 5) +
  geom_errorbarh(aes(xmax = estimate+std.error, xmin = estimate-std.error), height = 0, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Estimate", y = "Term") +
  scale_colour_brewer(palette = "Dark2") +
theme(
    panel.grid.major.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
) +ggsave("../Figures/model9_coefs.png")
```

```{r plot_model9, echo=FALSE, message=FALSE, warning=FALSE}
data_summary <- data_long %>%
  filter(HPP < 4) %>%
  group_by(HPP, trialType) %>%
  summarise(mean = mean(lookingTime))
data_long %>%
  filter(HPP < 4) %>%
  mutate(trialType = factor(trialType, c("Familiar", "Novel"))) %>%
  ggplot(aes(x = trialType, y = lookingTime)) +
  facet_wrap(~HPP, nrow = 1) +
  zero_margin +
  geom_violin(aes(fill = trialType), colour = NA, alpha = 0.5) +
  geom_line(aes(group = Participant), colour = "black", size = 0.25, alpha = 0.2) +
  geom_line(data = data_summary, aes(x = trialType, y = mean), size = 1, colour = "black") +
  stat_summary(fun.data = mean_se, geom = "pointrange", group = 1, size = 0.5) +
  stat_summary(fun.y = mean, geom = "line", group = 1, size = 0.5) +
  labs(x = "HPP", y = "Looking time (ms)") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
) +
  ggsave("../Figures/model9_interaction.png")

```

## How to interpret the distribution of absolute values of novelty preference scores?

To address this question, I ran two separate small-scale simulations based on the distribution of looking times in our existing data set. The general question is - what might the distribution of absolute novelty preferences scores look like, in general, even in the absence of underlying systematic individual differences in preference direction.

### Shuffled IDs

The first simulation I ran was simply re-shuffling infant ids across the observed looking times. On each iteration, I re-shuffled subject ids within the set of novel and familiar looking times for each study, while maintaining the proportion of infants with each number of HPP visits. This breaks the association between a specific looking time for familiar stimuli and a specific looking time for novel stimuli. 

Below is a plot of the distribution of **average** absolute NPS values after 1000 simulations. The observed absolute NPS values in the current study are plotted in black. As we can see, it seems like the observed absolute NPS values would not be surprising in the absence of systematic relationships between familiar and novel looking time stimuli, given the distribution of those values in the population in general.

```{r sim1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# simulation of noisy looking and using absolute values in preferential looking.

# shuffle infant ids across looking type

summary_stats_shuffling <- data.frame()

num_reps <- 1000

set.seed(100)

for (i in 1:num_reps) {
  d_long_shuffled <- data.frame()
  
  for (study_name in unique(data_long$Study)) {
    # break apart long data into a familiar and novel looking data frame
    familiar <- filter(data_long, trialType == "Familiar" & Study == study_name)
    novel    <- filter(data_long, trialType == "Novel"    & Study == study_name)
    
    # shuffle subject ids within each of these data frames
    # in other words, each subject becomes randomly associated with another looking time
    familiar_shuffled_order <- sample(1:nrow(familiar))
    familiar$baby_shuffled  <- familiar$Participant[familiar_shuffled_order]
    familiar$hpp_shuffled   <- familiar$HPP[familiar_shuffled_order]
    
    novel_shuffled_order <- sample(1:nrow(novel))
    novel$baby_shuffled  <- novel$Participant[novel_shuffled_order]
    novel$hpp_shuffled   <- novel$HPP[novel_shuffled_order]
    
    # recombine the two data sets
    temp            <- bind_rows(familiar, novel)
    d_long_shuffled <- bind_rows(d_long_shuffled, temp)
    
  }
  # compute novelty preference for shuffled baby labels
  data_shuffled <- d_long_shuffled %>%
    group_by(baby_shuffled, hpp_shuffled, Study) %>%
    summarize(
      novelty_pref          = lookingTime[trialType == "Novel"] - lookingTime[trialType == "Familiar"],
      novelty_pref_corr     = novelty_pref / (lookingTime[trialType == "Novel"] + lookingTime[trialType == "Familiar"]),
      novelty_pref_corr_abs = abs(novelty_pref_corr)
    )
  
  summary_shuffled <- data_shuffled %>%
    group_by(Study, hpp_shuffled) %>%
    summarize(
      N                       = sum(!is.na(novelty_pref_corr_abs)),
      M_abs_novelty_pref_corr = mean(novelty_pref_corr_abs),
      SD                      = sd(novelty_pref_corr_abs)
    ) %>%
    mutate(
      CI_low = M_abs_novelty_pref_corr - qt(0.975, df = N - 1)*SD/sqrt(N),
      CI_high = M_abs_novelty_pref_corr + qt(0.975, df = N-1)*SD/sqrt(N)
    )
  
  summary_shuffled$repetition <- i
  
  # add to overall data frame
  summary_stats_shuffling <- bind_rows(summary_stats_shuffling, summary_shuffled)
}

summary_absolute_preference <- data %>%
  mutate(novelty_pref_corr_abs = abs(novel_pref_corr)) %>%
  group_by(Study, HPP) %>%
  summarize(
    N                       = sum(!is.na(novelty_pref_corr_abs)),
    M_abs_novelty_pref_corr = mean(novelty_pref_corr_abs),
    SD                      = sd(novelty_pref_corr_abs)) %>%
  mutate(
    CI_low  = M_abs_novelty_pref_corr - qt(0.975, df = N-1)*SD/sqrt(N),
    CI_high = M_abs_novelty_pref_corr + qt(0.975, df = N-1)*SD/sqrt(N)
  )
```


```{r plot_sim1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# plot distribution of mean absolute novelty preference numbers, by study
ggplot(summary_stats_shuffling, aes(hpp_shuffled, M_abs_novelty_pref_corr))+
  facet_wrap(~Study) +
  geom_violin(aes(group = hpp_shuffled, fill = Study, colour = Study), alpha = 0.5, colour = NA) +
  geom_boxplot(aes(group = hpp_shuffled), width = 0.5, fill = "transparent", colour = "black") +
  geom_point(data = summary_absolute_preference, aes(x = HPP), color = "black", size = 4)+
  geom_errorbar(data = filter(summary_absolute_preference, N > 3),
                aes(x = HPP, ymin = CI_low, ymax = CI_high),
                color = "black", width = 0, size = 1) +
  scale_y_continuous(breaks = seq(0, 0.6, 0.1)) +
  scale_fill_brewer(palette = "Dark2") +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Number of HPP visits", y = "Mean Novelty Preference (%)\n(corrected for total looking time)") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
  ) +
  ggsave("../Figures/simulation1.png")

```

### Simulated data assuming no individual-level effect

Although I find the simulation above informative and fun, it's not entirely obvious to me that it's the strongest test of whether the absolute NPS values are meaningful, since it's not explicitly investigating the distribution of absolute values, **assuming absolutely no systematic preference**.

To investigate this question, I simulated new observations based on the variation in values in the current data set, assuming no significant preference for familiar and novel stimuli for each simulated individual and that the log looking time values are normally distributed.

Empirically, it appears from the plot below that the looking times themselves are not normally distributed, but the log-transformed values are somewhat consistent with a normal distribution.

```{r plot2_sim1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, message=F}

# analyze general mean and sd of the total data set, on corrected looking time values.
ggplot(data_long, aes(lookingTime, fill = trialType)) +
  facet_wrap(~ trialType + Study) +
  geom_histogram(aes(y = ..density..), binwidth = 500) +
  geom_density(color = "black", fill = "black", alpha = 0.4) +
  labs(x = "Looking time (ms)", y = "Density", title = "Raw looking time") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid       = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.background = element_rect(fill = "white", colour = "grey"),
    text             = element_text(colour = "black", size = 12),
    axis.text        = element_text(colour = "black"),
    legend.position  = "none"
  ) +
  ggplot(data_long, aes(log(lookingTime), fill = trialType)) +
  facet_wrap(~ trialType + Study) +
  geom_histogram(aes(y = ..density..), binwidth = 0.10) +
  geom_density(color = "black", fill = "black", alpha = 0.4) +
  labs(x = "Looking time", y = "Density", title = "Log-transformed looking time") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid       = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.background = element_rect(fill = "white", colour = "grey"),
    text             = element_text(colour = "black", size = 12),
    axis.text        = element_text(colour = "black"),
    legend.position  = "none"
  ) +
  plot_layout(nrow = 1) +
  ggsave("../Figures/distributions.png")

```

In this simulation, we regenerate a new dataset repeatedly from scratch, assuming no systematic familiarity/ novelty preference within each individual and that the looking time values are normally distributed with the same SD observed in the current study.

One tweak in this simulation is that I account for systematic relationships between familiar and novel looking times within a given subject by - for each subject - sampling from a distribution of average overall looking times, and then shifting the distribution of the looking time values based on that mean. Note that the familiar and novel looking times are sampled from the same distribution for each infant, i.e. each simulated participant has no preference for familiar vs. novel stimuli.

The plot below shows the distribution of average absolute NPS values for each study after 1000 simulations. As you can see, the level of average absolute NPS values observed in our dataset (plotted in black) would not be surprising in a situation where infants have **no preference** between novel and familiar stimuli.

```{r sim2, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# collapse across familiar/novel trials
# mean and sd
summarized_looking <- data_long %>%
  group_by(Study) %>%
  summarize(N              = sum(!is.na(novel_pref_corr)),
            M_looking      = mean(lookingTime),
            SD_looking     = sd(lookingTime),
            M_log_looking  = mean(log(lookingTime)),
            SD_log_looking = sd(log(lookingTime)))

average_looking_time_by_subject <- data %>%
  group_by(Study) %>%
  summarize(
    mean_avg_log_looking = mean(log(avg_looking_time)),
    sd_avg_log_looking   = sd(log(avg_looking_time))
    )


# simulate a new data set by drawing repeatedly from a normal distribution 
# based on log looking times in each study
# draw mean of distribution for each subject from distribution of average looking times, to account for "long lookers" vs. "short lookers"

num_repetitions <- 1000
random_d        <- data.frame()

for (i in 1:num_repetitions) {
  
  for (study_name in unique(summarized_looking$Study)) {
    n                <- filter(summarized_looking, Study == study_name)$N
    cur_sd           <- filter(summarized_looking, Study == study_name)$SD_log_looking
    avg_looking_mean <- filter(average_looking_time_by_subject, Study == study_name)$mean_avg_log_looking
    avg_looking_sd   <- filter(average_looking_time_by_subject, Study == study_name)$sd_avg_log_looking
    
    for (j in 1:n) {
      # sample mean from distribution of average looking times
      cur_mean <- rnorm(1, mean = avg_looking_mean, sd = avg_looking_sd)
      temp <- data.frame(
        Study                     = study_name, 
        repetition                = i,
        baby_id                   = paste("rep_", i, "_baby_", j, sep=""),
        log.lookingTime.novel    = rnorm(1, mean = cur_mean, sd = cur_sd),
        log.lookingTime.familiar = rnorm(1, mean = cur_mean, sd = cur_sd)
      )
    }
    
    # convert back log looking times to novelty preference score on looking times
    temp <- temp %>%
      mutate(novel = exp(log.lookingTime.novel), familiar = exp(log.lookingTime.familiar)) %>%
      mutate(
        novel_pref      = novel - familiar,
        novel_pref_corr = novel_pref / (familiar+novel))
    
    random_d <- bind_rows(random_d, temp)
    
  }
}

average_random_d <- random_d %>%
  group_by(repetition, Study) %>%
  summarize(
    mean_novelty_pref = mean(novel_pref_corr),
    mean_abs_novelty_pref = mean(abs(novel_pref_corr))
  )

average_abs_novelty_pref <- data %>%
  group_by(Study) %>%
  summarize(
    N                     = sum(!is.na(novel_pref_corr)),
    mean_abs_novelty_pref = mean(abs(novel_pref_corr)),
    SD                    = sd(abs(novel_pref_corr))
    ) %>%
  mutate(
    CI_low  = mean_abs_novelty_pref-qt(0.975, df = N-1)*SD/sqrt(N),
    CI_high = mean_abs_novelty_pref+qt(0.975, df= N-1)*SD/sqrt(N)
    )

```

```{r plot_sim2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(average_random_d, aes(Study, mean_abs_novelty_pref, fill = Study)) +
  geom_violin(size = 1, alpha = 0.5, colour = NA) +
  geom_boxplot(width = 0.5, fill = "transparent", colour = "black", width = 0.25, size = 0.75) +
  geom_point(data = average_abs_novelty_pref, color = "black", size = 4, width = 0.5, size = 0.75) +
  geom_errorbar(data = average_abs_novelty_pref, aes(ymin = CI_low, ymax = CI_high), color = "black", width = 0, size = 1) +
  labs(x = "Study", y = "Mean absolute novelty preference") +
  scale_y_continuous(breaks = seq(0, 0.6, 0.1)) +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid       = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background = element_rect(fill = "white", colour = "grey"),
    text             = element_text(colour = "black", size = 12),
    axis.text        = element_text(colour = "black"),
    legend.position  = "none"
  ) +
  ggsave("../Figures/simulation2.png")

```

