---
title: "SI2. Participants"
author:
    - Chiara Santolin*
    - Gonzalo García-Castro
    - Martin Zettersten
    - Nuria Sebastian-Galles
    - Jenny R. Saffran
date: "Updated: `r format(Sys.Date(), '%B %dth, %Y')`"
output:
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: flip_template.docx
bibliography: "../Flip.bib"
csl: "../apa.csl"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo    = TRUE,
	message = FALSE,
	warning = FALSE,
	include = TRUE,
	out.width = "100%"
)
options(knitr.kable.NA = '-', big.mark = ',') # to replace "NA" with "-" in knitr::kable

```

```{r prepare, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# load packages
library(tidyverse)    # tidyverse packages
library(magrittr)     # for using pipes
library(knitr)        # for formatting tables
library(english)      # for transforming numbers to words
library(here)         # for locating files

# import data
data <- read.csv(here("Data", "01_data-processed.csv"), header = TRUE) %>%
  mutate(Item = ifelse(Item==0, "Familiar", "Novel"))

```

# Participants

```{r participants, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# number of observations
n <- data %>% summarise(n = n())                                  
# number of participants
n_participants <- data %>% filter(Item == "Novel") %>% summarise(n = n()) 
# number of participants by study
n_study <- data %>%                                                           
  filter(Item == "Novel") %>%
  group_by(Study, Location) %>%
  summarise(n = n())
# number of participants by study and location
n_studyLocation <- data %>%
  filter(Item == "Novel") %>%
  group_by(Study, Location) %>%
  summarise(n = n())

```

We retrieved data from `r n_participants$n` participants: `r n_studyLocation$n[1]` from @saffran2003 (tested in Wisconsin), `r n_studyLocation$n[2]` from @saffran2008 (tested in Wisconsin), `r n_study$n[3]` from @santolin2019a (tested in Wisconsin), and `r n_studyLocation$n[4]` from the resplication study (@santolin2019; tested in Barcelona). Two datapoints were available for each participant: one for the mean looking time in "familiar" trials, and one for the mean looking time in "novel" trials. A total of `r n$n` datapoints were included in the analysis. Participants included in our analysis were those included in the final version of both studies.

## References

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent