---
title: "Methods"
output:
  pdf_document:
    fig_caption: yes
bibliography: ../Flip.bib
csl: ../apa.csl

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	include = TRUE
)
options(knitr.kable.NA = '-', big.mark = ',') # to replace "NA" with "-" in knitr::kable

```

```{r prepare, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# load packages
library(tidyverse)    # tidyverse packages
library(magrittr)     # for using pipes
library(lmerTest)     # for ANOVAs (computing Satterhwaite’s degrees of freedom)
library(optimx)       # for specifying lm4::lmer optimiser
library(predictmeans) # for checking LMM assumptions
library(effects)      # for analysing the interaction
library(car)          # for checking multicollinearity 
library(knitr)        # for formatting tables
library(broom.mixed)  # for tidy lm4::lmer output
library(patchwork)    # for grouping plots
library(english)      # to transform numbers to words

# import data
data <- read.csv("../Data/flip2.csv") %>%                                      # read in data
  mutate(avg_looking_time = (Familiar+Novel)/2) %>%                            # compute average looking time
  gather("trialType", "lookingTime", Familiar, Novel) %>%                      # make long format
  mutate(
    trialType       = factor(trialType, levels = c("Familiar", "Novel")),      # set "familiar" as baseline
    Study           = factor(Study, levels = c("Santolin", "Saffran & Wilson")), # make "Study" a factor
    HPP_center      = (HPP - mean(HPP)),                                        # center the HPP  
    trialType_center = ifelse(trialType=="Familiar",-0.5,0.5)
  )

```

## Participants

```{r participants, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

n              <- data %>% summarise(n = n())                                  # number of observations
n_participants <- data %>% filter(trialType == "Novel") %>% summarise(n = n()) # number of participants
n_study <- data %>%                                                            # number of participants by study
  filter(trialType == "Novel") %>%
  group_by(Study, Location) %>%
  summarise(n = n())
n_studyLocation <- data %>%                                                    # number of participants by study and location
  filter(trialType == "Novel") %>%
  group_by(Study, Location) %>%
  summarise(n = n())

```

We retrieved data from `r n_participants$n` participants: `r n_study$n[1]` from @santolin2019a (`r n_studyLocation$n[2]` tested in Wisconsin, the replication study [@santolin2019b; `r n_studyLocation$n[1]`] tested in Barcelona, and `r n_studyLocation$n[3]` from @saffran2003, tested in Wisconsin. Two datapoints were available for each participant: one for the mean looking time in "familiar" trials, and one for the mean looking time in "novel" trials. A total of `r n$n` datapoints were included in the analysis. Participants included in our analysis were those included in the final version of both studies.

## Studies

The two studies conducted by Santolin and colleages compared looking times in novel and familiar trials, and revealed that infants displayed a preference toward one of the trial types. The direction of the preference, however, was different for the two studies.

```{r plot_data, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.align="c", fig.cap="Looking time (ms) in familiar and novel trials split by Study and Location. Error bars indicate the standard error of the mean, respectively. Grey lines indicate participant level mean looking times."}

ggplot(data, aes(trialType, lookingTime, fill = Study))+
  facet_wrap(~ Study + Location) +
  stat_summary(fun.y = mean, geom = "bar", colour = NA) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", colour = "black", width = 0.25, size = 0.75) +
  labs(x = "Trial type", y = "Looking time (ms)") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none" 
  )
```

## Data analysis

We fit a model predicting looking time ($LT$) from the fixed effects trial type ($TrialType$, centered; Familiar = -0.5 and Novel = 0.5), the number of HeadTurn Preference Procedure experiments completed by infants ($HPP$), and their interaction ($TrialType \times HPP$) as fixed effects. *Familiar* trials were set as the baseline. Participant ($Participant$) and study ($Study$) were included as random effects. Following Barr et al. (2013), we fit a model with the maximal random effects structure that allowed the model to converge. The maximal random effects structure included a by-participant and by-study random intercepts, and a random slope of $HPP$ by $Study$. We subsequently pruned the random effects structure until convergence was achieved. The final model included by-participant and by-study random intercepts. The particular random effects structure chosen does not qualitatively impact the estimates and conclusions from the model

This model accounts for across-participants variability in overall looking time (i.e. some infants are long lookers, some are short lookers), and for across-studies differences in overall looking time, and allows the effect of $HPP$ to vary across studies. We specifiyied by-study random effects for two strong reasons. First, participants from different linguistic/cultural environments were included in both studies. This may have led to participants in one of the locations to looking longer in average than those from the other location. Second, in spite of their similarity both studies were not identical, which could also have led to differences in overall looking time.

```{r model, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
model <- lmer(
  lookingTime ~              # response variable
    trialType_center * HPP_center + # fixed effects ("*" means "include the interaction")
    (1 | Participant) +      # by-participant random intercept
    (1 + HPP_center|Study),  # by-study random intercept and HPP random slope
  data = data,               # indicate dataset
  REML = TRUE               # fit using REML
) 

#does not converge (singular boundary fit due to high correlation between by-study random intercept and by-study random slope for HPP)
#prune the random effects structure by removing the by-study random slope for HPP (similar results)

model <- lmer(
  lookingTime ~              # response variable
    trialType_center * HPP_center + # fixed effects ("*" means "include the interaction")
    (1 | Participant) +      # by-participant random intercept
    (1|Study),  # by-study random intercept and HPP random slope
  data = data,               # indicate dataset
  REML = TRUE               # fit using REML
) 

summary <- summary(model) %$% # extract coefficients from model
  coefficients %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  mutate(term = c("(Intercept)", # rename terms to join datasets 
                  "trialType_center",
                  "HPP_center",
                  "trialType_center:HPP_center")) %>% 
  select(., term, coefficient = Estimate)

confidence.intervals <- confint.merMod( # calculate confidence intervals
  model,  
  method = "Wald",
  level = 0.95
) %>%
  as.data.frame() %>%
  rownames_to_column(var = "term") %>%
  slice(4:7) %>%
  mutate(term = c("(Intercept)" , # rename terms to join datasets 
                  "trialType_center",
                  "HPP_center",
                  "trialType_center:HPP_center")) 

anova <- anova(model, type = 3, ddf = "Satterthwaite") %>% # perform type III ANOVA using Satterthwaite's approximation to degrees of freedom
  tidy() %>%
  right_join(., summary, by = "term") %>% # join outcome with coefficients
  left_join(., confidence.intervals, by = "term") %>% # join outcome with confidence intervals
  mutate(ci1 = round(`2.5 %`, 2),
         ci2 = round(`97.5 %`, 2)) %>%
  unite(col = "ci", ci1, ci2, sep = ", ") %>% # make a string with the lower and upper CI
  select(-c(`2.5 %`, `97.5 %`)) %>%
  select(term, coefficient, ci, statistic, DenDF, p.value)

#alternative with Kenward Rogers degrees of freedom (virtually identical, some odd outcomes with intercept)
# anova <- Anova(model, type="III",test="F") %>% # perform type III ANOVA using Kenward Rogers approximation to degrees of freedom
#   tidy() %>%
#   right_join(., summary, by = "term") %>% # join outcome with coefficients
#   left_join(., confidence.intervals, by = "term") %>% # join outcome with confidence intervals
#   mutate(ci1 = round(`2.5 %`, 2),
#          ci2 = round(`97.5 %`, 2)) %>%
#   unite(col = "ci", ci1, ci2, sep = ", ") %>% # make a string with the lower and upper CI
#   select(-c(`2.5 %`, `97.5 %`)) %>%
#   select(term, coefficient, ci, statistic, Df.res, p.value)


anova %>%
  mutate(term = c("Intercept", "Trial type", "HPP", "Trial type * HPP")) %>%
  kable(digits = 3, col.names = c("Term", "Coefficient", "95% CI", "t", "df", "p"), align = "c") 

```

We found a statistically significant interaction term, *t*(`r round(anova$DenDF[4], 2)`) = `r round(anova$statistic[4], 2)`, *p* = `r round(anova$p.value[4], 3)`, 95% CI = [`r anova$ci[4]`], suggesting that the effect of trial type on looking time was influenced by the number of HPP experiments each participant participated in. The interaction shows that experience with a higher number of HPP experiments is associated with a stronger novelty preference.

```{r plot_model, echo=FALSE, message=FALSE, warning=FALSE, fig.align='c', fig.cap="Looking times of familiar and novel trials, split by number of HeadTurn Preference Procedure experiments. Black points and error bars represent the group-level mean looking time and SEM, respectively. Grey lines represent participant-level mean looking time."}
sample_sizes <- data %>%              # calculate mean looking time by HPP and trialType
  group_by(HPP, trialType) %>%
  summarise(n = n()) %>%
  mutate(n = paste0("n = ", n))
data %>%
  ggplot(aes(x = trialType, y = lookingTime, fill = trialType)) +
  facet_wrap(~HPP, nrow = 1, switch = "x") +
  stat_summary(fun.y = mean, geom = "bar", size = 0.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25, size = 0.75) +
  geom_text(data = sample_sizes, aes(x = trialType, y = 250, label = n), colour = "white") +
  labs(x = "HPP", y = "Looking time (ms)") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none",
    strip.placement = "outside" 
)
```


```{r interaction, echo=FALSE, message=FALSE, warning=FALSE, fig.align="c", fig.cap="(A) Coefficients of the fixed effects. Dots, whiskers, and shaded boxes represent the estimated coefficients, standard error, and 95 percent Wald confidence intervals, respectively. (B) Predicted looking times plotted against HPP, split by study."}

coefficients <- tidy(model) %>%                        # extract coefficients
  filter(effect == 'fixed', term != "(Intercept)") %>% # get coefficients of interest
  mutate(term = c("trialType_center",                         # rename terms to join datasets
                  "HPP_center",
                  "trialType_center:HPP_center")) %>% 
  left_join(., confidence.intervals, by = "term") %>%  # join outcome with confidence intervals         
  mutate(term = c("Trial type",                        # rename terms for joining datasets
                  "HPP",
                  "Trial type * HPP"),
         lower.se = estimate-std.error,
         upper.se = estimate+std.error) %>%
  rename(ci_lower = `2.5 %`, ci_upper = `97.5 %`)

effects <- effect(term = "trialType_center*HPP_center", mod = model) %>%
  as.data.frame()

#uncenter x-variables in model predictions to create a more interpretable plot
effects_uncentered <- effect(term = "trialType_center*HPP_center", mod = model) %>%
  as.data.frame() %>%
  filter(trialType_center %in% c(-0.5, 0.5)) %>%
  mutate(
    trialType = ifelse(trialType_center==0.5,"Novel","Familiar"),
    HPP = HPP_center + mean(data$HPP)
  )

# plot coefficients
ggplot(data = coefficients, aes(term, estimate, colour = term, fill = term)) +
  geom_linerange(aes(x = term, ymin = ci_lower, ymax = ci_upper), alpha = 0.7, size = 10) +
  geom_point(size = 5, colour = "black") +
  geom_errorbar(aes(ymax = estimate+std.error, ymin = estimate-std.error), height = 0, size = 1.5, width = 0, colour = "black") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  labs(x = "Term", y = "coefficient") +
  scale_colour_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
  ) +
  coord_flip() +
  
  # add interaction plot 
  ggplot(effects_uncentered, aes(x = HPP, y = fit, fill = trialType, colour = trialType)) +
  geom_smooth(method="lm",size = 1,fullrange=T) +
  labs(x = "HPP", y = "Looking time (ms)", colour = "Trial type", fill = "Trial type") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
  ) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = "A")+
  scale_x_continuous(breaks=seq(0,5,1), limits=c(1,5))+
  scale_y_continuous(breaks=seq(5000,8000,500),limits=c(5000,7750))

```


# Appendices

### Appendix 1: Linear Mixed Model

Following @barr2013 guidelines, we initially specified a maximal random structure, including random intercepts for $Participant$ and $Study$, and random slopes for $TrialType$, $HPP$, and $TrialType \times HPP$, by $Participant$ and $Study$. Due to lack of convergence, the random structure was simplified until the model fit was no longer singular. The code and formula of the mode are presented below:

<center>`lookingTime ~ TrialType * HPP + (1 | Participant) + (1 | Study) + (1 + HPP | Study)`</center>

$$ \begin{aligned}LT_{ips} = \\ &\beta_{0} + Participant_{0p} + Study_{0s} + \\ &(\beta_{1} + Study_{s}) HPP_{i} + \\ &\beta_{2}TrialType_{i} + \\ &(\beta_{3} + Study_s) HPP \times TrialType_{i} + \\ &e_{ips},e_{ips} \sim N(0, \sigma^2), Participant_{0p} \sim N(0, \tau_{00}^2), Study_{0p} \sim N(0, &\omega_{00}^2)\end{aligned}$$

Where:

* $LT_{ips}$ is the looking time in trial $i$, in participant $p$ from study $s$.
* $\beta_{0}$ is the fixed intercept (the grand mean of all looking times from all trials).
* $Participant_{0p}$ is the by-participant random intercept (the overall looking time of participant $p$, assumed to have been sampled from a normal distribution with mean 0 and variance $\tau_{00}^2$.
* $Study_{0s}$ is the by-study random intercept (the overall looking time in study $s$, assumed to have been sampled from a normal distribution with mean 0 and variance $\omega_{00}^2$.
* $\beta_{1}$ is the coefficient of the fixed effect of the $HPP$ predictor.
* $\beta_{2}$ is the coefficient of the fixed effect of the $TrialType$ predictor.
* $\beta_{3}$ is the coefficient of the fixed effect of the $TrialType \times HPP$ interaction.
* $e_{ips}$ is the error of the model in trial $i$, of participant $p$ from study $s$, assumed to be normally distributed with mean 0 and variance $\sigma^2$.

Before fitting the model, we group mean-centered the $HPP$ predictor.The model was fit using the `lme4` R package [@lme4]. We used the `lmerTest` package [@lmerTest] to compute *p*-values using the Satterthwaite's approximation to degrees of freedom [@satterthwaite1946a]. 


```{r posterior_predicted_simulation, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
iqrvec      <- sapply(simulate(model, 1000), IQR)
obsval      <- IQR(data$lookingTime)
post.pred.p <- mean(obsval >= c(obsval, iqrvec))

```

Posterior predicted simulations [@gelman2006] indicated that the coefficients obtained by our model are plausible: The observed inter-quartile range (IQR) of looking times lied within the IQR obtained from 1000 simulated datasets generated from the model in more than 95% of the simulations (*p* = `r round(post.pred.p, 3)`).

### Appendix 2: Checking assumptions of the Linear Mixed Model

Residuals seem to approximate a normal distribution, though the some scores in the @santolin2019a and @santolin2019b studies seem to be somewhat shifted.

```{r normality, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

ggplot(data = fortify(model), aes(sample = .resid)) +
  geom_qq(alpha = 0.3) +
  geom_qq_line(colour = "black", size = 1) +
  labs(x = "Theoretical", y = "Sample", colour = "Trial") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
  ) +
  ggplot(data = fortify(model), aes(sample = .resid, colour = Study)) +
  facet_wrap(~Study) +
  geom_qq(alpha = 0.3) +
  geom_qq_line(size = 1) +
  labs(x = "Theoretical", y = "Sample", colour = "Study") +
  scale_colour_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
  ) +
  ggplot(data = fortify(model), aes(x = .resid)) +
  geom_density(alpha = 0.5, fill = "black") +
  labs(x = "Residuals", y = "Density", colour = "Study") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
  ) +
  ggplot(data = fortify(model), aes(x = .resid, fill = Study)) +
  facet_wrap(~Study) +
  geom_density(alpha = 0.5) +
  labs(x = "Residuals", y = "Density", colour = "Study") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.x = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "none"
  ) +
  plot_layout(nrow = 2, ncol = 2)
  

```


```{r linearity_homoskedasticity, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
data.frame(
  studentised_residual = rstudent(model),
  fitted = fitted(model),
  study = data$Study
) %>%
ggplot(aes(fitted, studentised_residual, colour = study, fill = study)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", fullrange = TRUE) +
  labs(x = "Predicted", y = "Studentised residuals", colour = "Study", fill = "Study") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    panel.grid.major.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.minor.y = element_line(colour = "grey", size = 0.25, linetype = "dotted"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.background   = element_rect(fill = "white", colour = "grey"),
    text               = element_text(colour = "black", size = 12),
    axis.text          = element_text(colour = "black"),
    legend.position    = "right"
  )
  
```

We observed little evidence of multicollinearity in the predictors we included in the model:

```{r multicollinearity, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
vif(model) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(vif = ".") %>%
  mutate(term = c("Trial type", "HPP", "Trial type * HPP"),
         tolerance = 1/vif) %>%
  kable(digits = 2, col.names = c("", "VIF", "Tolerance"), align = "c")
``` 

## References